---
title: "_T_ chimera: Batch correction"
author: "C. Guibentif, J. Griffiths et al. Chimera"
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
---


```{r setup, message=FALSE}
library(Matrix)
library(scran)
library(Rtsne)
library(irlba)
library(cowplot)
library(biomaRt)

source("/nfs/research1/marioni/jonny/chimera-t/scripts/core_functions.R")
load_data(remove_doublets = TRUE, remove_stripped = TRUE, remove_pool2 = TRUE)

nPC = 50


```

In this script, we perform batch correction on our data.


#Batch correction

For batch correction, we employ the scran function `fastMNN`, which performs batch correction in the manner of `mnnCorrect`, but in the PC-space, and much faster. Critically, this is a composition-aware batch-correction strategy that should not be affected by the lack of e.g. extraembryonic tissue in the Tomato+ samples (as the injected stem cells do not contribute to these lineages). We correct within each timepoint only.

##Total correction

```{r mnncorrect-1}

hvgs_E7.5 = getHVGs(scater::normalize(sce[,meta$stage == "E7.5"]))
hvgs_E8.5 = getHVGs(scater::normalize(sce[,meta$stage == "E8.5"]))

tab = table(meta$sample[meta$stage == "E7.5"])

#perform batch correction within each genotype, then across the genotypes
correct1 = doBatchCorrect(counts = logcounts(scater::normalize(sce[hvgs_E7.5, meta$stage == "E7.5"])), 
                          timepoints = meta$tomato[meta$stage == "E7.5"], 
                          samples = meta$sample[meta$stage == "E7.5"], 
                          timepoint_order = as.logical(c("TRUE", "FALSE")), 
                          sample_order = as.numeric(names(tab[order(tab, decreasing = TRUE)])))

```

```{r mnncorrect-2}

tab = table(meta$sample[meta$stage == "E8.5"])

correct2 = doBatchCorrect(counts = logcounts(scater::normalize(sce[hvgs_E8.5, meta$stage == "E8.5"])), 
                          timepoints = meta$tomato[meta$stage == "E8.5"], 
                          samples = meta$sample[meta$stage == "E8.5"], 
                          timepoint_order = as.logical(c("TRUE", "FALSE")), 
                          sample_order = as.numeric(names(tab[order(tab, decreasing = TRUE)])))


```

```{r mnncorrect-join}

corrected = list("E7.5" = correct1,
                 "E8.5" = correct2)

saveRDS(corrected, file = "/nfs/research1/marioni/jonny/chimera-t/data/corrected_pcas.rds")



```

A t-SNE visualisation of our cells, pre- and post-correction, is shown in Figure \@ref(fig:compare-all).

```{r compare-all, fig.height = 10, fig.wide = TRUE, fig.cap = "t-SNE of cells before and after correction. Cells in red colours are Tomato+ (injected), cells in grey colours are Tomato- (embryo). Different shades of colour indicate difference samples."}

base_7.5 = prcomp_irlba(t(logcounts(scater::normalize(sce[hvgs_E7.5, meta$stage == "E7.5"]))), n = nPC)$x
base_8.5 = prcomp_irlba(t(logcounts(scater::normalize(sce[hvgs_E8.5, meta$stage == "E8.5"]))), n = nPC)$x

tsne_pre_7.5 = Rtsne(base_7.5, pca = FALSE)$Y
tsne_post_7.5 = Rtsne(corrected$E7.5, pca = FALSE)$Y

tsne_pre_8.5 = Rtsne(base_8.5, pca = FALSE)$Y
tsne_post_8.5 = Rtsne(corrected$E8.5, pca = FALSE)$Y

ro1 = sample(nrow(base_7.5), nrow(base_7.5))
ro2 = sample(nrow(base_8.5), nrow(base_8.5))

p1 = ggplot(as.data.frame(tsne_pre_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E7.5"][ro1]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = sample_colours, name = "Sample") +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E7.5, pre-correction") +
  labs(x= "tSNE1", y= "tSNE2")

p2 = ggplot(as.data.frame(tsne_post_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E7.5"][ro1]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = sample_colours, name = "Sample") +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E7.5, post-correction") +
  labs(x= "tSNE1", y= "tSNE2")

p3 = ggplot(as.data.frame(tsne_pre_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E8.5"][ro2]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = sample_colours, name = "Sample") +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E8.5, pre-correction") +
  labs(x= "tSNE1", y= "tSNE2")

p4 = ggplot(as.data.frame(tsne_post_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$sample[meta$stage == "E8.5"][ro2]))) +
  geom_point(size = 0.4) +
  scale_colour_manual(values = sample_colours, name = "Sample") +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E8.5, post-correction") +
  labs(x= "tSNE1", y= "tSNE2")

plot_grid(p1, p2, p3, p4)

```


# Celltype-coloured tSNE

Figure \@ref(fig:celltype) shows the same plots, but coloured by the mapped celltype (see the mapping script for details).

```{r celltype, fig.wide = FALSE, fig.height=12, fig.cap = "t-SNEs, coloured by celltype."}


tsne_7.5 = Rtsne(corrected[["E7.5"]], pca = FALSE)$Y
tsne_8.5 = Rtsne(corrected[["E8.5"]], pca = FALSE)$Y

ro1 = sample(sum(meta$stage == "E7.5"), sum(meta$stage == "E7.5"))
ro2 = sample(sum(meta$stage == "E8.5"), sum(meta$stage == "E8.5"))

plegend = ggplot(as.data.frame(tsne_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$celltype.mapped[meta$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours, drop = FALSE, name = "") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())  +
  guides(col = guide_legend(override.aes = list(size = 5), ncol = 3))

p1 = ggplot(as.data.frame(tsne_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$celltype.mapped[meta$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours, name = "") +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E7.5") +
  labs(x = "tSNE1", y = "tSNE2")


p2 = ggplot(as.data.frame(tsne_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$celltype.mapped[meta$stage == "E8.5"][ro2], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point() +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E8.5") +
  labs(x = "tSNE1", y = "tSNE2")


plot_grid(p1, get_legend(plegend), p2, ncol = 1, rel_heights = c(1,0.6,1))


```

# Celltype-coloured UMAP

Finally, we generate UMAP coordinates of the batch-corrected data.

```{r generate-umap}

umap_7.5 = getUMAP(corrected[["E7.5"]])


umap_8.5 = getUMAP(corrected[["E8.5"]])

```

```{r plot-umap, fig.height = 14, fig.wide = FALSE}

p1 = ggplot(as.data.frame(umap_7.5)[ro1,], aes(x = V1, y = V2, col = factor(meta$celltype.mapped[meta$stage == "E7.5"][ro1], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point(size = 0.3) +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E7.5") +
  labs(x = "UMAP1", y = "UMAP2")


p2 = ggplot(as.data.frame(umap_8.5)[ro2,], aes(x = V1, y = V2, col = factor(meta$celltype.mapped[meta$stage == "E8.5"][ro2], levels = names(celltype_colours), ordered = TRUE))) +
  geom_point(size = 0.3) +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("E8.5") +
  labs(x = "UMAP1", y = "UMAP2")

plot_grid(p1, get_legend(plegend), p2, ncol = 1, rel_heights = c(1,0.6,1))

ggsave(p1, file = "/nfs/research1/marioni/jonny/chimera-t/scripts/batch_correct/umap_7.5.pdf",
       width = 5, height = 5)
ggsave(p2, file = "/nfs/research1/marioni/jonny/chimera-t/scripts/batch_correct/umap_8.5.pdf",
       width = 5, height = 5)

saveRDS(list("E7.5" = umap_7.5,
             "E8.5" = umap_8.5),
        file = "/nfs/research1/marioni/jonny/chimera-t/data/umaps.rds")

```

And below, the cells are split by their Tomato status.

```{r plot-umap-split, fig.height = 12, fig.wide = TRUE}

meta$tomatoText = ifelse(meta$tomato, "Tom+", "Tom-")

p1 = ggplot(data = meta[meta$stage == "E7.5",], 
            mapping = aes(x = umap_7.5[ro1,1], 
                          y = umap_7.5[ro1,2], 
                          col = factor(celltype.mapped[ro1], 
                                       levels = names(celltype_colours), 
                                       ordered = TRUE))) +
  geom_point(size = 0.3) +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E7.5") +
  facet_wrap(~tomatoText[ro1])

p2 = ggplot(data = meta[meta$stage == "E8.5",], 
            mapping = aes(x = umap_8.5[ro2,1], 
                          y = umap_8.5[ro2,2], 
                          col = factor(celltype.mapped[ro2], 
                                       levels = names(celltype_colours), 
                                       ordered = TRUE))) +
  geom_point(size = 0.3) +
  scale_colour_manual(values = celltype_colours) +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  ggtitle("E8.5") +
  facet_wrap(~tomatoText[ro2])

plot_grid(p1, p2, ncol = 1)

ggsave(p1, file = "/nfs/research1/marioni/jonny/chimera-t/scripts/batch_correct/umap_7.5_split.pdf",
       width = 10, height = 5)
ggsave(p2, file = "/nfs/research1/marioni/jonny/chimera-t/scripts/batch_correct/umap_8.5_split.pdf",
       width = 10, height = 5)


```

## Clone differences

These chimeras contain two differenct clones of the T knockout (i.e., different disruptions of the T genes). These are shown in Figure \@ref(fig:clones) for all samples.


```{r clones, fig.wide = TRUE, fig.height = 16, fig.cap = "Location of different clones in the UMAPs"}

meta$clone = sapply(meta$sample,
                    switch,
                    "B6KO",
                    "B6HOST",
                    "B6KO",
                    "B6HOST",
                    "B6KO",
                    "B6HOST",
                    "B6KO",
                    "B6HOST",
                    "C6KO",
                    "C6HOST",
                    "B6KO",
                    "B6HOST",
                    "C6KO",
                    "C6HOST",
                    "C6KO",
                    "C6HOST")

meta_8.5= meta[meta$stage == "E8.5",]
plots_8.5 = lapply(unique(meta_8.5$clone), function(x){
  ggplot() +
    geom_point(mapping = aes(x = umap_8.5[,1], 
                             y = umap_8.5[,2]), 
               col = "lightgrey") +
    geom_point(mapping = aes(x = umap_8.5[meta_8.5$clone==x,1], 
                             y = umap_8.5[meta_8.5$clone==x,2], 
                             col = meta_8.5$celltype.mapped[meta_8.5$clone==x])) +
    scale_color_manual(values = celltype_colours) +
    theme(legend.position = "none",
          axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = "UMAP1", y = "UMAP2") +
    ggtitle(paste0(x, " (E8.5)"))
})

meta_7.5= meta[meta$stage == "E7.5",]
plots_7.5 = lapply(unique(meta_7.5$clone), function(x){
  ggplot() +
    geom_point(mapping = aes(x = umap_7.5[,1], 
                             y = umap_7.5[,2]), 
               col = "lightgrey") +
    geom_point(mapping = aes(x = umap_7.5[meta_7.5$clone==x,1], 
                             y = umap_7.5[meta_7.5$clone==x,2], 
                             fill = meta_7.5$celltype.mapped[meta_7.5$clone==x]),
    pch = 21, col = "black") +
    scale_fill_manual(values = celltype_colours) +
    theme(legend.position = "none",
          axis.text = element_blank(),
          axis.ticks = element_blank()) +
    labs(x = "UMAP1", y = "UMAP2") +
    ggtitle(paste0(x, " (E7.5)"))
})

grid = plot_grid(plotlist= c(plots_7.5, plots_8.5), ncol = 2)

grid

save_plot(grid, file = "/nfs/research1/marioni/jonny/chimera-t/scripts/batch_correct/clone.pdf", base_width = 8, base_height = 16)

```
#Session Info
```{r sessinf}
sessionInfo()
```
