---
title: "_T_ chimera: Swapped molecule removal"
author: "C. Guibentif, J. Griffiths et al. _T_ Chimera"
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
---


In this script, we exclude molecules that are likely to have derived from barcode swapping. In short, we identify molecules that share UMI, cell barcode, and aligned gene between samples that were sequenced together. These molecules are extremely unlikely to have arisen due to chance, and are instead far more likely to derive from barcode swapping. For more information on this phenomenon, please read [our paper](https://www.nature.com/articles/s41467-018-05083-x). 

Figure \@ref(fig:plot-ratios) shows the fraction of molecules that were excluded, on a sample-by-sample basis. Note that the WT and T chimeras were sequenced together, but we only visualise the T samples that belong to this chimera dataset. 
<!-- While swapped molecule estimates are calculated for all samples in this script, it is the wild-type chimera counterpart that saves the data to disk, except for samples 11 to 14, which were multiplexed only with each other. -->

```{r do-batch1, message = FALSE}

library(DropletUtils)
library(ggplot2)
library(cowplot)
library(Matrix)

mol_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 1:6, "/outs/molecule_info.h5"),
             paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 1:2, "/outs/molecule_info.h5"))
out_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 1:6, "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"),
            paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 1:2, "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"))
bc_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 1:6, "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"),
           paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 1:2, "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"))
gene_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 1:6, "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"),
             paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 1:2, "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"))


unswapped = swappedDrops(mol_loc, get.swapped = TRUE)

for(i in 1:length(mol_loc)){
  null_holder = writeMM(unswapped$cleaned[[i]], file = out_loc[i])
  write.table(colnames(unswapped$cleaned[[i]]), file = bc_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
  write.table(rownames(unswapped$cleaned[[i]]), file = gene_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
}

unswapped$swapped = unswapped$swapped[7:8]
unswapped$cleaned = unswapped$cleaned[7:8]

ratios = sapply(1:length(unswapped$cleaned), function(i){
  sum(unswapped$swapped[[i]])/(sum(unswapped$cleaned[[i]]) + sum(unswapped$swapped[[i]]))
})

```

```{r do-batch2, message = FALSE}


mol_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 7:10, "/outs/molecule_info.h5"),
             paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 3:10, "/outs/molecule_info.h5"))
out_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 7:10, "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"),
            paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 3:10, "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"))
bc_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 7:10, "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"),
           paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 3:10, "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"))
gene_loc = c(paste0("/nfs/research1/marioni/jonny/chimera-wt/cellranger_out/sample_", 7:10, "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"),
             paste0("/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 3:10, "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"))


unswapped = swappedDrops(mol_loc, get.swapped = TRUE)

for(i in 1:length(mol_loc)){
  null_holder = writeMM(unswapped$cleaned[[i]], file = out_loc[i])
  write.table(colnames(unswapped$cleaned[[i]]), file = bc_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
  write.table(rownames(unswapped$cleaned[[i]]), file = gene_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
}

unswapped$swapped = unswapped$swapped[5:12]
unswapped$cleaned = unswapped$cleaned[5:12]

ratios_2 = sapply(1:length(unswapped$cleaned), function(i){
  sum(unswapped$swapped[[i]])/(sum(unswapped$cleaned[[i]]) + sum(unswapped$swapped[[i]]))
})
ratios = c(ratios, ratios_2)

```

```{r do-batch3, message = FALSE}


mol_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  11:14, 
  "/outs/molecule_info.h5"
  )
out_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  11:14, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"
  )
bc_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  11:14, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"
  )
gene_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  11:14, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"
  )


unswapped = swappedDrops(mol_loc, get.swapped = TRUE)

for(i in 1:length(mol_loc)){
  null_holder = writeMM(unswapped$cleaned[[i]], file = out_loc[i])
  write.table(colnames(unswapped$cleaned[[i]]), file = bc_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
  write.table(rownames(unswapped$cleaned[[i]]), file = gene_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
}

ratios_3 = sapply(1:length(unswapped$cleaned), function(i){
  sum(unswapped$swapped[[i]])/(sum(unswapped$cleaned[[i]]) + sum(unswapped$swapped[[i]]))
})
ratios = c(ratios, ratios_3)

```

```{r do-batch4, message = FALSE}

#Note that these data were multiplexed with some samples from another researcher
#we don't have access to their data, so will only be able to remove swapped molecules between our two samples.
mol_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  15:16, 
  "/outs/molecule_info.h5"
  )
out_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  15:16, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/matrix_unswapped.mtx"
  )
bc_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  15:16, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/barcodes_unswapped.tsv"
  )
gene_loc = paste0(
  "/nfs/research1/marioni/jonny/chimera-t/cellranger_out/sample_", 
  15:16, 
  "/outs/raw_gene_bc_matrices/mm10_tomato/genes_unswapped.tsv"
  )


unswapped = swappedDrops(mol_loc, get.swapped = TRUE)

for(i in 1:length(mol_loc)){
  null_holder = writeMM(unswapped$cleaned[[i]], file = out_loc[i])
  write.table(colnames(unswapped$cleaned[[i]]), file = bc_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
  write.table(rownames(unswapped$cleaned[[i]]), file = gene_loc[i], col.names = FALSE, row.names = FALSE, quote = FALSE)
}

ratios_4 = sapply(1:length(unswapped$cleaned), function(i){
  sum(unswapped$swapped[[i]])/(sum(unswapped$cleaned[[i]]) + sum(unswapped$swapped[[i]]))
})
ratios = c(ratios, ratios_4)

```

```{r plot-ratios, fig.cap = "Fraction of molecules removed due to barcode-swapping correction."}

ggplot(data.frame(ratios = ratios, sample = 1:length(ratios)), aes(x = sample, y = ratios)) +
  geom_bar(stat = "identity", fill = "grey20") +
  labs(x = "Sample", y = "Fraction of molecules excluded")

```